---
title: "Colias_RPanda"
author: "Mariana P Braga"
format: 
  html:
    code-fold: show
    code-summary: "Show the code"
editor: visual
---

## Diversification rate shifts in Colias

Packages

```{r}
#| label: "packages"
#| message: FALSE

# install branch from RPanda repo
#install_github("hmorlon/PANDA", ref = "clade.shift.model")

# load packages
library(ape)
library(RPANDA)
library(RColorBrewer)

# library(phytools)
# library(phangorn)
# library(parallel)
# library(ParallelLogger)
# library(RColorBrewer)
# library(devtools)
# library(pspline)

```

Load the data

```{r}

tree <- readRDS("tree_Colias.rds") 
groups <- readRDS("groups.rds")

```

Shift combinations

```{r}
#| fig-height: 10
#| fig-width: 7

sfrac <- get.sampling.fractions(tree, groups, plot = TRUE, cex = 0.5)
node_comb <- get.comb.shift(tree, groups, sfrac)

par(mfrow=c(2,2))
for(i in 1:4){
  plot.phylo.comb(phylo = tree, 
                data = groups, 
                sampling.fractions = sfrac, 
                combi = node_comb[i], 
                edge.width = 2, 
                label.offset = 0.3, 
                lad = F, 
                cex = 0.4)
  axisPhylo()
}
dev.off()
```

Find best combination of models

```{r}
shifts <- shift.estimates(phylo = tree,
                          data = groups,
                          sampling.fractions = sfrac,
                          comb.shift = node_comb,
                          models = c("BCST","BVAR",
                                     "BCST_DCST","BVAR_DVAR",
                                     "BVAR_DCST","BCST_DVAR"),
                          backbone.option = "crown.shift",
                          multi.backbone = "all",
                          Ncores = 4)

str(shifts, max.level = 1)
head(shifts$total)
shifts$backbones$`75/`

```

Rates and diversity across time

```{r}
rates <- div.rates(phylo = tree,
                   shift.res = shifts,
                   time.interval = 0.5,
                   combi = 1, 
                   part = "all")
rates

paleodiversity <- paleodiv(phylo = tree,
                           data = groups,
                           sampling.fractions = sfrac,
                           time.interval = 0.5,
                           shift.res = shifts,
                           split.div = TRUE)
paleodiversity
```

Plot shifts, diversification rates and paleodiversity over time

```{r}
#| label: "big plot"
#| fig-height: 9
#| fig-width: 10

# Define the layout of plot with the matrix mat 
mat <- matrix(c(1,1,2,3), nrow = 2, ncol = 2)

layout(mat) 
par(mar = c(6,4,3,4)) 

subclade_colors <- brewer.pal(8, "Dark2")[1:4] 
group_colors <- c(subclade_colors, "black")

# Phylogeny with shifts 

plot.phylo.comb(phylo = tree, data = groups, sampling.fractions = sfrac, shift.res = shifts, combi = 1, label.offset = 0.1, cex = 0.7, lad = F) 
axisPhylo() 
mtext("Time (Ma)", side = 1, line = 2, cex = 1, at = 17)

# Rates through time 

crown_age <- max(node.depth.edgelength(tree))
time <- -c(crown_age, seq(floor(crown_age), 0, by = -0.5))

plot(time, rates[[1]][1,], type = "l", col = group_colors[1], las = 1, xlab = "Time (Myrs)", ylab = "Rates (Events/ Lineages / Myr)", main = "Diversification rates")
lines(time, rates[[length(rates)]][1,], type = "l", col = "black") 
legend("topright", legend = c("Backbone", "After shift"), col = c("black", group_colors[1]), lty = 1)

# Paleodiversity dynamics 
plot(time, paleodiversity["75",],
     las = 1, ylim = c(0, 80),
     col = group_colors[1],
     main = "Paleodiversity dynamics of Colias",
     xlab = "Time (Myrs)", ylab = "Number of species", type = "l") 
lines(time, paleodiversity["backbone",], col = "black")
legend("topleft", legend = c("Backbone", "After shift"), col = c("black", group_colors[1]), lty = 1)
```


## Model adequacy

To test whether the selected model is adequate, we can simulate data with that model and then compare them to the empirical data. If the model is adequate, it will produce data similar to the observed empirical data.

```{r}
#| label: "adequacy"



```


